Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-11 paper/main.tex
update some comments, include initial SEPIA cov fn info

Let $S$ represent the true (but unknown) underlying power matter spectrum for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e., 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence of the
low resolution runs (as visualized in Figure \ref{fig:plot_data}).  For
this, we embrace variations of the Mat\'ern kernel \citep{stein1999interpolation},
tailored to each scenario (more on this in later subsections).
We consider $\Sigma_\epsilon$ as known, given domain-matter expertise. (right)?
\textcolor{red}{(Do we actually tailor this?  I thought it was consistent across the board.)}

@@

These results come from Bayes' Theorem, with details of this derivation provided 
in \ref{sec:apdx_SgivenY}. With all posterior draws $S^{(t)}$, we compute
the average of these to obtain our posterior mean for a particular cosmology.
We denote our inference procedure for the hierarchical model of Eq.~\ref{eq:dgphm} as
``DGP.HM'' given its deep Gaussian process foundation with additional hierarchical modeling (HM)
structure to incorporate multiple functional outputs.
\textcolor{red}{(How do you get UQ though?)}

@@

% This is old: switched to SEPIA
% We assume a powered exponential kernel for the Gaussian process and employ 
% \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
% \citep{macdonald2015gpfit}. 
In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
GPs to model the principal components' weights. For direct comparison alongside
\cite{moran2023mira}, we employ a Python implementation of this approach called 
SEPIA \citep{gattiker2020sepia}. \textcolor{red}{Please note: currently show GPfit 
predictions for Mira-Titan Data, NOT SEPIA. The CAMB results are based on SEPIA.}
That is, the $i$th vector of PC weights $w_i(\psi)$, 
$i \in \{1,\ldots,p_\eta\}$, will be modeled as a zero mean GP: 
$w_i(\psi) \sim GP(0, \sigma^2R)$, with

\begin{equation*}
    R_{ij} = \prod_{k=1}^{p_\psi}\exp\left\{-10^{\beta_k}|x_{ik}-x_{jk}|^\alpha\right\}
\end{equation*}

\textcolor{red}{We fix $\alpha=1.95$ and estimate each $\beta_k$ with $k \in \{1,...,p_\psi\}$. 
The above with $R_{ij}$ is based on GPfit; will update this to reflect SEPIA settings later.}
Using these trained GPs from the PCs, we predict the spectra for unobserved cosmologies.

>>>>

Let $S$ represent the true (but unknown) underlying power matter spectrum for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e., 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence of the
low resolution runs (as visualized in Figure \ref{fig:plot_data}).  For
this, we embrace variations of the Mat\'ern kernel \citep{stein1999interpolation},
tailored to each scenario (more on this in later subsections).
We consider $\Sigma_\varepsilon$ as known, given domain-matter expertise.
\textcolor{red}{(Here, we are fixing $\Sigma_\varepsilon$ upon estimating the Matern 
parameters and utilizing the precision estimates (domain-matter expertise from 
log-log regression model)?  The estimate of $\Sigma_\varepsilon$ is tailored for
each cosmology, as the Matern estimates are computed uniquely for each.)}

@@

These results come from Bayes' Theorem, with details of this derivation provided 
in \ref{sec:apdx_SgivenY}. With all posterior draws $S^{(t)}$, we compute
the average of these to obtain our posterior mean for a particular cosmology. Additionally,
we quantify uncertainty by finding the 2.5th and 97.5th percentiles of the posterior 
draws at each index to obtain a 95\% credible interval.
We denote our inference procedure for the hierarchical model of Eq.~\ref{eq:dgphm} as
``DGP.HM'' given its deep Gaussian process foundation with additional hierarchical modeling (HM)
structure to incorporate multiple functional outputs.

@@

% This is old: switched to SEPIA
% We assume a powered exponential kernel for the Gaussian process and employ 
% \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
% \citep{macdonald2015gpfit}. 
In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
GPs to model the principal components' weights. For direct comparison alongside
\cite{moran2023mira}, we employ a Python implementation of this approach called 
SEPIA \citep{gattiker2020sepia}. That is, the $i$th vector of PC weights $w_i(\psi)$, 
$i \in \{1,\ldots,p_\eta\}$, is modeled as a zero-mean Gaussian process with a 
squared exponential (i.e., Gaussian) covariance function:
\[
w_i(\psi) \sim GP\left(0,\ \frac{1}{\lambda_{z_i}} R^{(i)} + \frac{1}{\lambda_{s_i}} I \right),
\]
with covariance function
\[
R^{(i)}_{uv} = \exp\left(-\sum_{k=1}^{p_\psi} \beta_{ik} (x_{uk} - x_{vk})^2 \right),
\]
where $\beta_{ik}$ is the inverse squared lengthscale for the $k$th input dimension, $\lambda_{z_i}$ is the marginal precision for $w_i(\psi)$, and $\lambda_{s_i}$ is the nugget precision (set to zero if no nugget is used).

Using these trained GPs from the PCs, we predict the spectra for unobserved cosmologies.
